Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2019-08-13 journals/04-lime_algorithm_and_proposed_methods/04-lime_algorithm_and_proposed_methods.Rmd
worked on procedure in paper

# The Data

# Draft of LIME Procedure

@@

>>>>

# LIME Procedure

### Original Paper

*Note: Most of the text in this section is copied from the original paper.*

| Notation | Original Paper Definition | My Interpretation (in terms of a data table) |
| -------- | ------------------------- | ------------------------- |
| $x\in\mathbb{R}^d$ | Original representation of an instance being explained | Vector of observed features for one observation e.g. $(x_1, x_2,...,x_d)$ where $\textbf{X}$ is an $n$ by $d$ matrix | 
| $f:\mathbb{R}^d→\mathbb{R}$ | The model being explained...In classification, $f(x)$ is the probability (or a binary indicator) that $x$ belongs to a certain class. | This is a random forest in our case. |
| $x'\in\{0,1\}^{d'}$ | A binary vector for interpretable representation (where interpretable explanations are defined as "use a representation that is understandable by humans") | Varies based on the simulation method used - note that $1\le d' \le d$ since it is possible that feature selection may be performed to choose a smaller set of variables than originally in $x$ (One example: vector of indicator variables that indicate whether the observed value falls in the same bin as the instance being explained e.g. $x'_i = I[x_i\in \mbox{same variable } i \mbox{ bin as the instance being explained}]$) | In our example, this is the random forest model. |
|$G$ | A class of potentially interpretable models, such as linear models, decision trees, or falling rule lists |  <span style="color:blue"> I'm not sure what is considered $G$ in the R package - the fitting of the final ridge regression model or the ridge regression model used for feature selection? </span>|
| $g$ | An explanation as a model...can be readily presented to the user with visual or textual artifacts...The domain of $g$ is $\{0,1\}^{d′}$, i.e. $g$ acts over absence/presence of the interpretable components | This could be a ridge regression model with the form $f(x) \sim x'_1 + x'_2 + \cdots + x'_{d'}$ |
| $\Omega(g)$ | A measure of complexity of the explanation $g\in G$. For example, for decision trees $\Omega(g)$ may be the depth of the tree, while for linear models, $\Omega(g)$ may be the number of non-zero weights. | Seemingly, this is based on how the weights are assigned to the data points. |
| $\pi_x(z)$ | A proximity measure between an instance $z$ to $x$, so as to define locality around $x$. | This could be the Gower distance. |
| $\mathcal{L}(f,g,\pi_x)$ | A measure of how unfaithful $g$ is in approximating $f$ in the locality defined by $\pi_x$. | <span style="color:blue"> Does the LIME R package ever consider this? </span> |

**Goal of LIME**: In order to ensure both interpretability and local fidelity, we must minimize $\mathcal{L}(f,g,\pi_x)$ while having $\Omega(g)$ below enough to be interpretable by humans. The explanation produced by LIME is obtained by the following:
  $$\xi(x) = \underset{g∈G}{\arg\min} \ \mathcal{L}(f,g,\pi_x) + \Omega(g)$$
This  formulation  can  be  used  with  different  explanation families $G$, fidelity functions $\mathcal{L}$, and complexity measures $\Omega$. Here we focus on sparse linear models as explanations, and on performing the search using perturbations.

**Procedure**: We want to minimize the locality-aware loss $\mathcal{L}(f,g, \pi_x)$ without making any assumptions about $f$, since we want the explainer to be model-agnostic.

1. In order to learn the local behavior of $f$ as the interpretable inputs vary, we approximate $\mathcal{L}(f,g,\pi_x)$ by drawing samples, weighted by $\pi_x$. 
2. We sample instances around $x'$ by drawing nonzero elements of $x'$ uniformly at random (where the number of such draws is also uniformly sampled).
3. Given a perturbed sample $z'\in\{0,1\}^{d′}$ (which contains a fraction of the nonzero elements of $x'$), we recover the sample in the original representation $z\in\mathbb{R}^d$ and obtain $f(z)$, which is used as a label for the explanation model. 
4. Given this data set $\mathcal{Z}$ of perturbed samples with the associated labels, we optimize $\mathcal{L}(f,g,\pi_x) + \Omega(g)$ to get an explanation $\xi(x)$.

It is worth noting that our method is fairly robust to sampling noise since the samples are weighted by $\pi_x$.

### R Package (Draft)

@@

### The Data

### 2019-08-13 papers/lime_diagnostics/visual-diagnostics-of-a-model-explainer.Rnw
worked on procedure in paper

\kg{***Need to clean up the notation.} Let $\textbf{X}$ be an $n$ by $p$ data matrix with $p$ features and $n$ observations, and let $x=\left(\begin{array}{cccc}x_1 & x_2 & \hdots & x_p\end{array}\right)\in\mathbb{R}^p$ be the observation of interest. Furthermore, let $\textbf{y}$ be a vector of length $n$ of response values and $y\in\{0, 1\}$ be the observed response value associated with $x$. Suppose that $f$ is a classification model where $f:\mathbb{R}^p\rightarrow[0,1]$ that is applied to $\textbf{X}$ and $\textbf{y}$. Let the predictions made by $f$ applied to $\textbf{X}$ be denoted as $\hat{\textbf{y}}$. It is of interest to explain the prediction made by $f$ when $f$ is applied to $x$. Note that $f(x)$ is equal to the probability that $y=1$. Given this setup, the LIME procedure is as follows.

@@

\item Apply a transformation $T$ to $\textbf{X}^*$ that results in an interpretable representation. The transformation that is applied will depend on the simulation method used. Section \ref{r package} will discuss the transformations used by the LIME R package. Let $T(\textbf{X}^*)=\textbf{X}'$. Furthermore, apply $T$ to $x$ to obtain $T(x)=x'$. \noindent\kg{***I need to check if the predictions are made before or after the transformation.}

\item Apply $f$ to $\textbf{X}'$ to obtain a vector of predictions $\hat{\textbf{y}}'$ of length $m$. Let the prediction made by $f$ when applied to the transformed case of interested be denoted as $f(x')=y'$.

>>>>

\kg{***Need to clean up the notation. Should I use the same notation as the original paper?} Let $\textbf{X}$ be an $n$ by $p$ data matrix with $p$ features and $n$ observations, and let $x=\left(\begin{array}{cccc}x_1 & x_2 & \hdots & x_p\end{array}\right)\in\mathbb{R}^p$ be the observation of interest. Furthermore, let $\textbf{y}$ be a vector of length $n$ of response values and $y\in\{0, 1\}$ be the observed response value associated with $x$. Suppose that $f$ is a classification model where $f:\mathbb{R}^p\rightarrow[0,1]$ that is applied to $\textbf{X}$ and $\textbf{y}$. Let the predictions made by $f$ applied to $\textbf{X}$ be denoted as $\hat{\textbf{y}}$. It is of interest to explain the prediction made by $f$ when $f$ is applied to $x$. Note that $f(x)$ is equal to the probability that $y=1$. Given this setup, the LIME procedure is as follows.

@@

\item Apply $f$ to $\textbf{X}^*$ to obtain a vector of predictions $\hat{\textbf{y}}'$ of length $m$. Let the prediction made by $f$ when applied to the transformed case of interested be denoted as $f(x')=\hat{y}'$.

\item Apply a transformation $T$ to $\textbf{X}^*$ that results in an interpretable representation. The transformation that is applied will depend on the simulation method used. Section \ref{r package} will discuss the transformations used by the LIME R package. Let $T(\textbf{X}^*)=\textbf{X}'$. Furthermore, apply $T$ to $x$ to obtain $T(x)=x'$.
